---
title: "HDS 5310 Final Project"
output:
  html_document:
    df_print: paged
---

-------------------------------
##### PROLOG #######
# Final Project
# Group 2
# Maternal Mortality Rate in the United States
# HDS 5310
# Deena Rakshitha Borelli, Megan Johnson, Monyya Sree Kommineni, Matt Mueller
# 05/08/2025
--------------------------------

```{r, include=FALSE}
# Load libraries
library(tidyverse)
library(tableone)
library(knitr)       
library(kableExtra)
library(car)
```


Load in the merged dataset
```{r}
data <- read.csv("final_MMR_dataset.txt")

# Convert categorical variables to factors
data$Year <- factor(data$Year)
data$Race <- factor(data$Race)
data$Urban <- factor(data$Urban,
                     # Reorder the levels to go from most urban to most rural
                     levels=c(
                       "Large Central Metro",
                       "Large Fringe Metro",
                       "Medium Metro",
                       "Small Metro",
                       "Micropolitan (Nonmetro)",
                       "NonCore (Nonmetro)"
                     ))

# Look at summary of data
summary(data)
```

#### Calculate descriptive statistics ####
```{r}
# Use CreateTableOne to generate the descriptive statistics table
tableOne <- CreateTableOne(data=data,
                           vars=c("Year", "Race", "Urban", "Deaths", "Births", "MMR"),
                           factorVars = c("Year", "Race", "Urban"))

# Put the table in a data frame with median and IQR for the continuous variables
table_df <- as.data.frame(print(tableOne, nonnormal = c("Deaths", "Births", "MMR"), quote=FALSE))

# Get the row names for the table
table_df <- data.frame(Variable = rownames(table_df), 
                               table_df, 
                               row.names = NULL)

# Use kable to make table
table_df %>%
  kable(format="html",
        caption="MMR data descriptive statistics",
        row.names=FALSE) %>%
  kable_styling()
```


#### Visualize data ####
```{r}
# Create histogram to show distribution of MMR
ggplot(data, aes(x=MMR)) +
  geom_histogram(binwidth=5, fill="lightblue", color="black") +
  theme_bw() +
  labs(title="Distribution of MMR",
       x="MMR (per 100,000 births)")
```

Interpretation:
 - The histogram appears right-skewed. It is clearly not normally-distributed.
 - MMR is <60 for most groups, with a few exceptions.


```{r}
# Create box plot of MMR by race
ggplot(data, aes(x=Race, y=MMR)) +
  geom_boxplot() +
  theme_bw() +
  labs(title="MMR by Race",
       x="Race",
       y="MMR (per 100,000 births)")
```

Interpretation:
 - The box plots by race show MMR is higher in Black mothers than it is for Asian or White.
 - The difference is notable, as the IQR for Black/African American does not overlap with the IQRs of the other groups.

```{r}
# Create box plot of MMR by Urban
ggplot(data, aes(x=Urban, y=MMR)) +
  geom_boxplot() +
  theme_bw() +
  labs(title="MMR by Urbanization Category",
       x="Urbanization classification",
       y="MMR (per 100,000 births)") +
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

Interpretation:
 - The difference between different urbanization classifications is less notable than for race.
 - The median MMR does increase as the groups become more rural (from left to right in this graph).
 - This trend suggests greater mortality in mothers who reside in less urban places, possibly due to being further from medical care.

```{r}
# Create scatter plot of MMR and year, colored by race
ggplot(data, aes(x=Year, y=MMR, color=Race)) +
  geom_point() +
  theme_bw() +
  labs(title="MMR over time, colored by Race",
       x="Year",
       y="MMR (per 100,000 births)")
```

Interpretation:
 - There does not appear to be an obvious difference between the years.
 - There appears to be a possible increase in MMR in 2020 and 2021. 
 - The coloring by Race group clearly shows again that MMR is higher in Black mothers, in all years 2018-2023.

```{r}
# Create scatter plot of MMR and year, colored by urbanization
ggplot(data, aes(x=Year, y=MMR, color=Urban)) +
  geom_point() +
  theme_bw() +
  labs(title="MMR over time, colored by Urbanization Category",
       x="Year",
       y="MMR (per 100,000 births)")
```

Interpretation:
 - No clear trend is observed by coloring by Urban.
 - This visualization does not suggest anything about the MMR and Urban over this time period.

#### Statistical tests ####
```{r, include=FALSE}
library(dplyr)
library(ggplot2)
library(car)
library(stats)
library(ggpubr)
```

#c. Statistical Tests, Assumptions & Alternatives

ANOVA for Race
```{r}
# Assumptions check for ANOVA
model_anova <- aov(MMR ~ Race, data = data)
shapiro.test(residuals(model_anova))           # Normality
leveneTest(MMR ~ Race, data = data)              # Equal variance (from car package)
# If assumptions are violated
kruskal.test(MMR ~ Race, data = data)
```

Interpretation:
Shapiro-Wilk:
 - Null hypothesis for Shapiro-Wilk test is that the distribution of residuals is normal.
 - The Shapiro-Wilk test p-value of 1.54e-6 is small enough to reject the null hypothesis.
 - The distribution of residuals is not normal, so the normality of residuals assumption for ANOVA is not met.
Levene:
 - Null hypothesis for Levene test is that the variance is equal across groups.
 - The Levene test p-value of 0.0002662 is small enough to reject the null hypothesis.
 - The variances are not equal across groups, so the equal variance assumption is not met.
 
Assumptions were violated, so the alternative test (Kruskal-Wallis) will be used instead of ANOVA.
Kruskal-Wallis results:
 - Null hypothesis for Kruskal-Wallis test is that MMR distribution is the same across Race groups.
 - The Kruskal-Wallis test p-value of 8.452e-13 is much smaller than 0.05, so the null hypothesis is rejected.
 - We conclude that the median MMR is significantly different between at least two of the Race groups.
 - Post-hoc test is warranted to determine which groups are significantly different.



ANOVA for Urban
```{r}
# Assumptions check for ANOVA
model_anova <- aov(MMR ~ Urban, data = data)
shapiro.test(residuals(model_anova))           # Normality
leveneTest(MMR ~ Urban, data = data)              # Equal variance (from car package)
# If assumptions are violated
kruskal.test(MMR ~ Urban, data = data)
```

Interpretation:
Shapiro-Wilk:
 - Null hypothesis for Shapiro-Wilk test is that the distribution of residuals is normal.
 - The Shapiro-Wilk test p-value of 0.005389 is small enough to reject the null hypothesis.
 - The distribution of residuals is not normal, so the normality of residuals assumption for ANOVA is not met.
Levene:
 - Null hypothesis for Levene test is that the variance is equal across groups.
 - The Levene test p-value of 0.01762 is small enough to reject the null hypothesis.
 - The variances are not equal across groups, so the equal variance assumption is not met.
 
Assumptions were violated, so the alternative test (Kruskal-Wallis) will be used instead of ANOVA.
Kruskal-Wallis results:
 - Null hypothesis for Kruskal-Wallis test is that MMR distribution is the same across Urban groups.
 - The Kruskal-Wallis test p-value of 0.02313 is smaller than 0.05, so the null hypothesis is rejected.
 - We conclude that the median MMR is significantly different between at least two of the Urban groups.
 - Post-hoc test is warranted to determine which groups are significantly different.


Post-hoc for Race
```{r}
# Perform pairwise Wilcoxon rank sum test (Mann-Whitney U)
pairwise.wilcox.test(data$MMR, data$Race, 
                     p.adjust.method = "BH",  # Benjamini-Hochberg (FDR) correction
                     exact = FALSE)          # Set to FALSE for larger samples

```

Interpretation:
 - MMR differed significantly between Black and White (p=5.1e-12) and Black and Asian (p=5.7e-05) groups.
 - MMR did not differ significantly between White and Asian groups (p=0.12).

Post-hoc for Urban
```{r}
# Perform pairwise Wilcoxon rank sum test (Mann-Whitney U)
pairwise.wilcox.test(data$MMR, data$Urban, 
                     p.adjust.method = "BH",  # Benjamini-Hochberg (FDR) correction
                     exact = FALSE)          # Set to FALSE for larger samples

```

Interpretation:
 - Even though the Kruskal-Wallis test showed significance, pairwise comparisons between Urban groups are not significant.
 - The differences between the nonmetro groups and the large metro groups are the closest to being significant


Chi-Squared test to investigate associate between Race and Urban
```{r}
# Create contingency table
table_result <- table(data$Race, data$Urban)

# Run Chi-squared test
chisq_result <- chisq.test(table_result)

# View result
print(chisq_result)

```

Interpretation:
 - The p-value for the Chi-squared test is 0.2864, which is not significant.
 - A warning is giving, suggesting that the assumptions for Chi-squared may not be met.


Check chi-squared assumption
```{r}
chisq_result$expected

```

Interpretation:
 - Many expected values are <5, so the chi-squared assumptions are not met.
 - Should use Fisher's test as an alternative test.


```{r}
# If expected frequencies < 5
fisher.test(table_result)
```

Interpretation:
 - The Fisher's Exact Test result was not statistically significant (p=0.511).
 - The distribution of racial groups was not significantly different across urbanization categories in this data.
 - Racial disparities in MMR are not explained by geographic distribution alone.


Look at standardized residuals
```{r}
chisq_result$stdres

```

Interpretation: 
 - Even though Chi-squared assumptions were violated, we can still look at standardized residuals to get a general sense of which values deviated most from the expected.
 - The largest residual was Asian/ Large Central Metro (residual=2.52, so more Asians than expected in Large Central Metro areas).
 - No other pairs (Race/Urban) stand out.

#d. Post-hocs & Planned Comparisons

```{r}
# Create new binary grouping variable
data$UrbanGroup <- ifelse(grepl("Metro", data$Urban), "Urban", "Rural")

# Make it a factor
data$UrbanGroup <- factor(data$UrbanGroup)

```

```{r}
table(data$UrbanGroup)

```


UrbanGroup t-test
```{r}
# Independent t-test
t_test_result <- t.test(MMR ~ UrbanGroup, data = data, var.equal = FALSE)
print(t_test_result)

# 95% Confidence Interval
t_test_result$conf.int
```

Interpretation:
 - UrbanGroup variable compares metro and non-metro areas.
 - Null hypothesis for the Welch's t-test is that the mean MMR is the same in Urban and Rural groups.
 - The significant p-value (0.008708) allows us to reject the null hypothesis.
 - The 95% confidence interval does not include 0.
 - We conclude that MMR does differ between urban and rural areas.


#e. Measuring effect size if there is a statistical significance
```{r}
library(effectsize)
cohens_d(MMR ~ UrbanGroup, data = data)
```

Interpretation:
 - Cohen's d of 0.86 is a large effect size (>0.8).
 - The confidence interval does not include 0, suggesting that the effect is meaningful.
 - We conclude that there is a large difference in MMR between urban and rural settings.
 

#f. Specific model to run for the analysis 
```{r}
# General linear model
lm_model <- lm(MMR ~ Urban + Race, data = data)

# Summary of the model
summary(lm_model)
```

Interpretation:
 - We fit a linear model to look at the combined effect of Urban and Race on MMR.
 - The model explained about 76.79% (R-squared) of the variance in MMR (p-value < 2.2e-16).
 

```{r}
# ANOVA table for the linear model
anova(lm_model)

```

Interpretation:
 - ANOVA table for the linear model showed Urban and Race were both significant predictors of MMR.

#g. Additional appropriate analyses to perform
```{r}
# Base model with just Urban
model_base <- lm(MMR ~ Urban, data = data)

# Full model with Urban + Race + their interaction
model_full <- lm(MMR ~ Urban * Race, data = data)

```

#a. Linearity (scatter plot or partial residuals):
```{r}
# Basic scatterplot matrix (continuous predictors only)
pairs(data[, c("MMR", "Deaths")])

```

#b. Constant Variance (Homoscedasticity) using Breusch-Pagan test:
```{r}
library(lmtest)
bptest(model_full)
```

Interpretation:
 - Null hypothesis for Breusch-Pagan test is the variance of residuals is constant.
 - p-value is not significant (0.07722), so we do not reject the null hypothesis.
 - The assumption is not violated.
 
 
#d. Normally Distributed Residuals:
```{r}
residuals <- resid(model_full)

# Q-Q Plot
qqnorm(residuals)
qqline(residuals, col = "red", lwd = 2)


# Shapiro-Wilk test
shapiro.test(residuals)
```

Interpretation:
 - The QQ-plot suggests that the residuals are not normally distributed.
 - Null hypothesis for Shapiro-Wilk test is that residuals are normally distributed, the Shapiro-Wilk test is significant (p=4.828e-05).
 - Reject the null hypothesis and conclude same as from QQ-plot: residuals are not normally distributed, the normality assumption for linear regression is not met.

#Add Interaction Terms (Effect Modification)
```{r}
summary(model_full)
```

#compare models(partial f-test)
```{r}
anova(model_base, model_full)

```

Interpretation:
 - The interaction model is highly significant (p< 2.2e-16).
 - This means the interaction explains more variance in MMR than Urban does on its own.
 - The relationship between location and MMR differs across race groups.
 - However, the normality assumption for linear regression was violated, so we should be cautious about these conclusions.